---
title: 'RDBEScore: Using R for ICES Regional Database & Estimation System (RDBES)
  Data'
output:
  powerpoint_presentation:
    reference_doc: "./data-raw/styles.pptx"
    slide_level: 2
    keep_md: true
    incremental: true
  ioslides_presentation:
    incremental: true
  beamer_presentation:
    incremental: true
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", message = FALSE,
                      warning = FALSE)
```

## What RDBEScore does

An R package for working with data from the [ICES Regional Database & Estimation System (RDBES)](https://rdbes.ices.dk/#/):

- **Import and check** RDBES downloads
- **Prepare** the data: filter, clean, add zeros for species not found
- **Estimate** from the sampling design and **export** to InterCatch

For national data coordinators, data analysts and scientists.

## The workflow

| Step | Functions |
|------|-----------|
| Download | `icesRDBES::rdbes_download_data()` |
| Import | `createRDBESDataObject()` |
| Check | `print()`, `validateRDBESDataObject()` |
| Prepare | `filterRDBESDataObject()`, `findAndKillOrphans()`, `generateZerosUsingSL()` |
| Estimate | `doEstimationForAllStrata()`, `createDesignTree()`, `doEstimationOnDesignTree()`, `doRatioEstimationToTotal()` |
| Export | `exportEstimationResultsToInterCatchFormat()` |

## Installation

From the ICES r-universe:

```{r eval=FALSE, echo=TRUE}
install.packages(c("RDBEScore", "icesRDBES"),
  repos = c("https://ices-tools-prod.r-universe.dev",
            "https://cloud.r-project.org"))
```

The development version: `remotes::install_github("ices-tools-dev/RDBEScore@dev")`

```{r}
library(RDBEScore)
library(data.table)
```

## Downloading data

Download from the [RDBES web site](https://rdbes.ices.dk/#/), or from R with the [icesRDBES](https://ices-tools-prod.r-universe.dev/icesRDBES) package (needs RDBES access):

```{r eval=FALSE, echo=TRUE}
payload <- list(dataType = "CS", format = "CsvFilePerTable",
                hierarchies = list("H1"),
                csFilters = list(deYear = list("1965"),
                                 sdCountry = list("ZW")))
zipFile <- icesRDBES::rdbes_download_data(payload,
                                          dest_dir = "./data-raw/exampleData")
```

## Importing a download

`createRDBESDataObject()` reads the downloaded `.zip` file:

```{r}
importedH1 <- createRDBESDataObject(
  input = "./data-raw/exampleData/H1_2025_02_11.zip")
```

- One hierarchy at a time
- Also reads CL, CE, VD and SL downloads, folders of CSV files and lists of data frames

## Checking the data

Printing shows the tables, their rows and how the units were selected:

```{r}
importedH1
```

```{r}
validateRDBESDataObject(importedH1)  # silent when the object is valid
```

## Filtering

Filter by any field, then remove the records that lost their links:

```{r}
filtered <- filterRDBESDataObject(H1Example,
                                  fieldsToFilter = c("SDctry", "VDflgCtry"),
                                  valuesToFilter = c("ZW", "ZWVFA"))
filtered <- findAndKillOrphans(filtered)
```

- `getLinkedDataFromLevel()` finds the linked records in another table, e.g. the vessel that caught a fish

## Zeros and missing values

Before estimation the data must say what was looked for but not found:

- `generateZerosUsingSL()` adds zero records for species on the species list but not in the samples
- `generateNAsUsingSL()` marks species that were not looked for as missing

A species that was not found is a zero; a species that was not looked for is missing. The estimates need to know the difference.

## Two ways to estimate catch numbers at age

The same samples and the same landings (sprat, area 27.3.d.28.1, first quarter):

- **Simple**: pool all fish of the quarter and raise their age composition to the landings
- **Design-based**: follow the sampling design: weeks, vessels, landings, boxes, fish

The next slides compare the two.

## The sampling design

`createDesignTree()` lists every sampling unit, its parent and how it was selected (here both quarters; the comparison shows the first):

```{r}
x <- H8ExampleEE1
tree <- createDesignTree(x, "BV", bvAssess = "Age", ratio = c(SA = "weight"),
                         strictSampleSize = FALSE, varianceAsWR = "TE")
```

```{r, echo = FALSE}
knitr::kable(tree[, .(units = .N, selection = paste(unique(method),
                                                    collapse = ", ")),
                  by = .(table)])
```

## Simple estimate

```{r}
cs <- list(LEarea = "27.3.d.28.1", LEmetier6 = "OTM_SPF_16-31_0_0",
           TEstratumName = month.name[1:3], SAspeCodeFAO = "SPR")
cl <- list(CLarea = "27.3.d.28.1", CLquar = 1,
           CLmetier6 = "OTM_SPF_16-31_0_0", CLspecFAO = "SPR")
biol <- addCLtoLowerCS(H8ExampleEE1, cs, cl, lowerHierarchy = "C",
                       CLfields = "CLoffWeight")
simple <- doBVestimCANUM(biol, "sumCLoffWeight", classUnits = "Ageyear",
                         classBreaks = 0:8)
```

## Design-based estimate

```{r}
fish <- getLeafValuesBV(tree, x, classMeas = "Age")
age <- as.integer(sub("N|Age=", "", fish$var, fixed = TRUE))
fish[, var := paste0("N|Age=", ifelse(age >= 8, "8+", age))]  # plus group
est <- doEstimationOnDesignTree(tree,
         rbind(fish, getLeafValuesSA(tree, x, "SAsampWtMes")))
quarterOf <- function(nodes) paste0("Q", ceiling(match(
  getAncestorValue(tree, nodes, x, "TEstratumName"), month.name) / 3))
landings <- x$CL[CLspecFAO == "SPR" & CLmetier6 == "OTM_SPF_16-31_0_0",
                 .(L = 1000 * sum(CLoffWeight)), by = .(q = paste0("Q", CLquar))]
design <- doRatioEstimationToTotal(est, unique(fish$var), "SAsampWtMes",
                                   X = setNames(landings$L, landings$q),
                                   domainOf = quarterOf)
```

## Results side by side

```{r, echo = FALSE}
cmp <- merge(simple[, .(Age = as.character(Group), simple = totNum)],
             design[domain == "Q1",
                    .(Age = sub("N|Age=", "", var, fixed = TRUE),
                      design = est, se)], by = "Age")
cmp <- cmp[order(as.integer(sub("+", "", Age, fixed = TRUE)))]
knitr::kable(cmp[, .(Age,
                     `Simple (millions)` = simple / 1e6,
                     `Design-based (millions)` = design / 1e6,
                     `Difference (%)` = 100 * (design / simple - 1),
                     `CV (%)` = 100 * se / design)],
             digits = 1)
```

- Similar totals, different age compositions
- Only the design-based estimates have standard errors

## Why they differ: the weight of each sample

```{r, echo = FALSE}
raising <- function(nodes) {
  w <- rep(1, length(nodes))
  repeat {
    i <- match(nodes, tree$node)
    if (all(is.na(i))) break
    w <- w * fifelse(is.na(i), 1, tree$w[i])
    nodes <- tree$parent[i]
  }
  w
}
samples <- tree[table == "SA", .(node, SAid = as.numeric(sub("SA:", "", node)),
                                 W = raising(parent))]
samples <- x$SA[, .(SAid, sampled = SAsampWtMes, landing = SAtotalWtMes)][
  samples, on = "SAid"]
samples[, month := getAncestorValue(tree, samples$node, x, "TEstratumName")]
samples <- samples[month %in% month.name[1:3]]
samples <- samples[order(match(month, month.name), SAid)]
knitr::kable(samples[, .(Month = month, Sample = SAid,
                         `Share of the sampled weight (%)` =
                           100 * sampled / sum(sampled),
                         `Share of the landings represented (%)` =
                           100 * W * landing / sum(W * landing))],
             digits = 0)
```

- Simple: a sample counts by its weight
- Design-based: a sample counts by the landings it represents

## Estimates match textbook examples

Totals and standard errors for textbook examples (Lohr, *Sampling: Design and Analysis*, and the survey package) converted to RDBES data:

```{r, echo = FALSE}
# the example objects have SAsamp == "N" on all SA rows: mark them sampled
sampled <- function(x) {
  x$SA <- data.table::copy(x$SA)
  x$SA[, SAsamp := "Y"]
  x
}
rootOf <- function(res) res[res$recType == "DE", c("est.total", "se.total")]
apiclus2 <- sampled(Pckg_survey_apiclus2_H1)
book <- data.frame(
  Example = c("Lohr: agsrs", "Lohr: agstrat", "Lohr: gpa",
              "survey: apiclus2"),
  Design = c("simple random", "stratified", "one-stage cluster",
             "two-stage"),
  Published = c("916,927,110 (58,169,381)", "909,736,035 (50,417,248)",
                "1,130.4 (65.5)", "2,639,273 (799,638)"))
# Reference: Lohr, Sampling: Design and Analysis, and the survey package;
# all reproduced with survey::svytotal()
res <- rbind(
  rootOf(doEstimationForAllStrata(sampled(Pckg_SDAResources_agsrs_H1),
                                  "SAsampWtMes")),
  rootOf(doEstimationForAllStrata(sampled(Pckg_SDAResources_agstrat_H1),
                                  "SAsampWtMes")),
  rootOf(doEstimationForAllStrata(sampled(Pckg_SDAResources_gpa_H1),
                                  "SAsampWtMes")) / 100,
  rootOf(doEstimationForAllStrata(apiclus2, "SAsampWtMes", na = "zero",
                                  strictSampleSize = FALSE)))
# decimals as in the references
digits <- c(0, 0, 1, 0)
fmt <- function(z) mapply(formatC, z, digits = digits,
                          MoreArgs = list(format = "f", big.mark = ","))
book$RDBEScore <- paste0(fmt(res$est.total), " (", fmt(res$se.total), ")")
knitr::kable(book, col.names = c("Example", "Design", "Reference (SE)",
                                 "RDBEScore (SE)"))
```

## Exporting to InterCatch

- Put the estimates in a data frame with the InterCatch column names
- `exportEstimationResultsToInterCatchFormat()` writes the HI, SI and SD records
- The design-based estimates also fill the variance columns, e.g. `varNumLanded`

A worked example is in the vignette *03b Estimating population parameters: ratio estimator*.

## Learn more

- Documentation and vignettes: [ices-tools-dev.github.io/RDBEScore](https://ices-tools-dev.github.io/RDBEScore/), or `browseVignettes("RDBEScore")`
- Source code: [github.com/ices-tools-dev/RDBEScore](https://github.com/ices-tools-dev/RDBEScore)
- Problems and ideas: [open an issue](https://github.com/ices-tools-dev/RDBEScore/issues)
- Contributions are welcome: fork the repository and send a pull request
