---
title: "Manipulating RDBESDataObjects"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Manipulating RDBESDataObjects}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction 

The aim of this document is to illustrate some of the ways of manipulating RDBESDataObjects.

```{r}
library(RDBEScore)
```

## Prerequisites

First we'll load some example data from the RDBES and check it's valid.  It's a good idea to check your RDBESDataObjects are valid after any manipulations you perform. See how to import your own data in the vignette [Import RDBES data](raw-data-import.html)

To get an example Hierarchy 1 data:

```{r load}


# Hierarchy 1 demo data
myH1RawObject <- H1Example

# Number of rows in each non-null table
print(myH1RawObject)


validateRDBESDataObject(myH1RawObject, verbose = FALSE)



```


To get the **number of rows** in each non-null table you can do:
```{r}
unlist(summary(myH1RawObject)$rows)
```

or simply call the object:

```{r}
myH1RawObject #equivalent to print(summary(myH1RawObject))
```

To get an example Hierarchy 5 data:

```{r}
# Hierarchy 5 demo data
myH5RawObject <- H5Example

# Number of rows in each non-null table and hierarchy
print(myH5RawObject)

validateRDBESDataObject(myH5RawObject, verbose = FALSE)

```

## Sorting RDBESDataObject

If the data has a single hierarchy in the DE table a correct sort order is defined for it. You can use the **sort()** method for it. Printing the object sorts it automatically if possible.

```{r}
#before sorting
names(H8Example)
```

```{r}
#after sorting
names(sort(H8Example))
```

```{r}
#printing the summary
H8Example
```



## Combining RDBESDataObjects

RDBESDataObjects can be combined using the **combineRDBESDataObjects()** function. This might be required when different sampling schemes are used to collect on-shore and at-sea samples - it will often be required to combine all the data together before further analysis.

```{r combine}
myCombinedRawObject <- combineRDBESDataObjects(myH1RawObject,
                                               myH5RawObject)

# Number of rows in each non-null table and hierarchies
print(myCombinedRawObject)

validateRDBESDataObject(myCombinedRawObject, verbose = FALSE)
```

## Filtering RDBESDataObjects

RDBESDataObjects can be filtered using the **filterRDBESDataObject()** function - this allows the RDBESDataObject to be filtered by any field.  A typical use of filtering might be to extract all data collected in a particular ICES division.  


```{r filter}
myH1RawObject <- H1Example

# Number of rows in each non-null table
print(myH1RawObject)

myFields <- c("SDctry","VDctry","VDflgCtry","FTarvLoc")
myValues <- c("ZW","ZWBZH","ZWVFA" )

myFilteredObject <- filterRDBESDataObject(myH1RawObject,
                                         fieldsToFilter = myFields,
                                         valuesToFilter = myValues )

# Number of rows in each non-null table
unlist(summary(myFilteredObject)$rows)

validateRDBESDataObject(myFilteredObject, verbose = FALSE)
```
It is important to note that filtering is likely to result in "orphan" rows being produced so it is usual to also apply the **findAndKillOrphans()** function to the filtered data to remove these records.

```{r clean1}

myFilteredObjectNoOrphans <- 
  findAndKillOrphans(objectToCheck = myFilteredObject, verbose = FALSE)

validateRDBESDataObject(myFilteredObjectNoOrphans, verbose = FALSE)


```

You can also remove any records that are not linking to a row in the VesselDetails (VD) table using the **removeBrokenVesselLinks()** function.

```{r clean2}

myFields <- c("VDlenCat")
myValues <- c("18-<24" )
myFilteredObject <- filterRDBESDataObject(myFilteredObjectNoOrphans,
                                         fieldsToFilter = myFields,
                                         valuesToFilter = myValues )

myFilteredObjectValidVesselLinks <- removeBrokenVesselLinks(
                                  objectToCheck = myFilteredObject,
                                  verbose = FALSE)

validateRDBESDataObject(myFilteredObjectValidVesselLinks, verbose = FALSE)

```

Finally you can also remove any records that are not linking to an entry in the SpeciesListDetails (SL) table using the **removeBrokenSpeciesListLinks()** function.

```{r clean3}

myFields <- c("SLspeclistName")
myValues <- c("ZW_1965_SpeciesList" )
myFilteredObjectValidSpeciesLinks <- filterRDBESDataObject(myFilteredObjectValidVesselLinks,
                                         fieldsToFilter = myFields,
                                         valuesToFilter = myValues )

myFilteredObjectValidSpeciesLinks <- removeBrokenSpeciesListLinks(
                                  objectToCheck = myFilteredObjectValidSpeciesLinks,
                                  verbose = FALSE)

validateRDBESDataObject(myFilteredObjectValidSpeciesLinks, verbose = FALSE)

```



See also other package vignettes:

```{r echo=FALSE, results='asis'}
vignettes <- as.data.frame(browseVignettes(package = "RDBEScore")$RDBEScore)
for (i in seq_along(vignettes$PDF)) {
  cat(sprintf("- [%s](%s)\n", vignettes$Title[i], vignettes$PDF[i]))
}

#END
```

