---
title: "03a Estimating population parameters: unbiased estimator"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{03a Estimating population parameters: unbiased estimator}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


## Introduction 

The aim of this document is to illustrate how population totals can be estimated for a single sampling stage with the **estimMC(...)** function, and for all stages and strata of an *RDBESDataObject* with **doEstimationForAllStrata(...)**, **createDesignTree(...)** and **doEstimationOnDesignTree(...)**.

```{r}
library(RDBEScore)
library(data.table)
```

## Prerequisites

First we'll load some example data from the RDBES and check it's valid.  It's a good idea to check your RDBESDataObjects are valid after any manipulations you perform.

```{r load}
h1d <- H1Example
validateRDBESDataObject(h1d, verbose = FALSE)
h1d
```


## Single level Multiple Count Estimator

Let's first estimate the last level values for a single FMid so that estimation is done for one top level record 
```{r}
FMidSel <- "4033243"
BV <- h1d$BV[h1d$BV$FMid == FMidSel,]

estimMC(as.numeric(BV$BVvalueMeas),
        BV$BVnumSamp,
        BV$BVnumTotal, 
        method=unique(BV$BVselectMeth))
```

This **estimMC(...)** function is actually the core of the estimation functions running on multiple levels as well. For implementation details see: [Variance calculation functions using "Multiple count" estimator](https://github.com/ices-tools-dev/RDBEScore/issues/13) 

## Estimation workflow

Estimation is done directly on the *RDBESDataObject*, in two steps:

1. **createDesignTree(...)** collects the sampling units of all tables (DE, SD, the upper hierarchy tables, SA and, if needed, FM and BV) into one table, the *design tree*. Every unit points to its parent unit and carries the design variables that describe how it was selected from its parent: stratum, selection method, numbers total and sampled, selection and inclusion probabilities, and auxiliary variables. Sub-sampled SA rows point to their parent SA, and clustering adds a cluster stage.
2. **doEstimationOnDesignTree(...)** takes the values of the variable of interest for the lowest units and folds them up the tree. For each stratum of the child units of a parent unit the total is estimated with the generalised Horvitz-Thompson (multiple count) estimator, the same estimator as in **estimMC(...)**, and the totals of the strata are added up for the parent. This is repeated stage by stage up to the top.

The variance of each stratum estimate is the variance between the sampled units plus the variance coming from the lower stages. For designs without replacement the lower stage variances are needed: if one of them cannot be estimated (e.g. a single sampled unit in a stratum) the variance above it is NA. For designs with replacement they are not needed. The `varianceAsWR` argument of **createDesignTree(...)** approximates the variance of a stage as if the units had been selected with replacement; this is conservative for designs without replacement.

**doEstimationForAllStrata(...)** does both steps for variables of the SA table and returns the estimates for every stratum.

The following example uses a dataset that resembles in many ways real data at ICES area 27.3.d.28.1 from the Estonian Baltic Trawling fleet for 2022 sprat (*Sprattus sprattus*). The dataset has both total landings and commercial sampling data.

```{r}
H8ExampleEE1
```

The data has not gone through RDBES upload and download procedure and contains invalid data types but also may have some other validity problems. However the fields for estimation should be present and valid.

```{r eval=FALSE, echo=TRUE}
#to check data types
validateRDBESDataObject(H8ExampleEE1, checkDataTypes = TRUE, strict = FALSE)
```

The design tree shows the sampling stages: weeks (TE) are sampled within months, vessels (VS) within weeks, landings (LE) from vessels, and boxes of sprat (SA) from the landings.

```{r}
tree <- createDesignTree(H8ExampleEE1, "SA", strictSampleSize = FALSE)
tree[, .N, by = .(table, method, depth)]
```

In two weeks there are two vessels in the data although `VSnumSamp` is 1. **createDesignTree(...)** stops on this by default; here `strictSampleSize = FALSE` turns the error into a warning and both vessels are raised as if each had been the only one sampled.

```{r, results='asis'}
x <- H8ExampleEE1
#the SAsampWtMes is in grammes let's estimate in tonnes
x$SA$SAsampWtMes <- x$SA$SAsampWtMes / (1000 * 1000)
results <- doEstimationForAllStrata(x, "SAsampWtMes",
                                    strictSampleSize = FALSE,
                                    varianceAsWR = "TE")
#let's compare the results from 1st quarter with actual catches from Q1
CL <- H8ExampleEE1$CL
Q1months <- c("January", "February", "March")
Q1results <- results[results$recType == "TE" &
                     results$stratumName %in% Q1months, ]
cat("Total estimated SPR catch in Q1: ",
    round(sum(Q1results$est.total), 3),  "tonnes\n")
#CLoffWeight is in Kg, converting to tonnes
cat("Total reported  SPR catch in Q1: ",
    sum(CL[CL$CLquar == 1, "CLoffWeight"]) / 1000, "tonnes")
```

Most weeks have only one sampled vessel, so the variance within the weeks cannot be estimated and the unbiased variance of the month totals is NA. Above, `varianceAsWR = "TE"` approximates it by treating the weeks as sampled with replacement.

```{r}
#let's add the actual totals
Q1results$actualTotals <- c(February = 124.829, January = 119.108,
                            March = 152.273)[Q1results$stratumName]
```

So the total estimated catch (*est.total*) for the first quarter is near the total of the actual catch. The table below gives the estimates by month with standard errors (*se.total*).

```{r}
columns <- c("recType", "stratumName", "n.units", "actualTotals",
             "est.total", "se.total")
captionText <- paste("Estimation results for 1st quarter by months (strata).",
                     "Total estimated catch (est.total) and standard error",
                     "(se.total), and the actual total reported weights.")
knitr::kable(Q1results[columns], digits = 3, caption = captionText)
```

As you can see from the table for the first 2 months the estimates are in line with the reported landings, while for March there is a big discrepancy.

In the next sections we will use data from R packages [survey](https://CRAN.R-project.org/package=survey) and [SDAResources](https://CRAN.R-project.org/package=SDAResources) that are converted into the *RDBESDataObject* to demonstrate the estimation procedure.

### Estimation on Multiple Strata

From [SDAResources](https://CRAN.R-project.org/package=SDAResources) we use *agstrat* which is data from a stratified random sample of size 300 from the 1992 U.S. Census of Agriculture agpop data. The dataset is converted into *RDBESDataObject* **Pckg_SDAResources_agstrat_H1** so that Table VS and SA contain the core information of \code{data(agstrat)} used in Lohr examples 3.2 and 3.6. Table VS is stratified with VSstratumName set to *region*, and VSnumberSampled and VSnumberTotal set according to *agstrat*. Table SA contains the variable measured *acres92* in \code{SAtotalWeightMeasured} and \code{SAsampleWeightMeasured}.
Table DE, SD, FT and FO are for the most dummy tables inserted to meet RDBES model requirements. Their selection method is CENSUS, so they do not change the estimates.

```{r}
agstrat_H1 <- Pckg_SDAResources_agstrat_H1
agstrat_H1
```

`SAsamp` is "N" on every SA row of this example object although all values are present. Units that were selected but not sampled stop the estimation (see `nonResponse` in **createDesignTree(...)**), so we mark the samples as sampled.

```{r}
agstrat_H1$SA$SAsamp <- "Y"
res <- doEstimationForAllStrata(agstrat_H1, "SAsampWtMes")
# Get the estimated total and its standard error for the VS stratum "NC"
columns2Get <- c("est.total", "se.total")
round(unlist(res[res$recType == "VS" & res$stratumName == "NC", columns2Get]), 1)
# and for the whole population
round(unlist(res[res$recType == "DE", columns2Get]), 1)
```

These are the estimates of Lohr (Example 3.2): the total acreage in the North Central region is 316,731,380 acres (SE 16,977,399) and in the whole population 909,736,035 acres (SE 50,417,248).

### Two Level Estimation on Single Stratum

From [survey](https://CRAN.R-project.org/package=survey) we use apiclus2 which is the Academic Performance Index computed for all California schools based on standardised testing of students. The design is 2-stage cluster sampling with clusters of unequal sizes. An SRS of 40 districts is selected (psus) from the 757 districts in the population and then up to 5 schools (min 1) were selected from each district (ssus).

The dataset is converted into *RDBESDataObject* **Pckg_survey_apiclus2_H1** so that 1 DE row with DEstratumName == "Pckg_SDAResources_apiclus2_H1", 1 child SD row, 40 child rows in VS (the 40 districts), VSnumberTotal is 757, VSnumberSampled is 40, 126 child rows in FT (the 126 schools finally observed), each associated to its cluster (dname), FTnumberTotal is the number of schools in district, FTnumberSampled is 1...5 schools sampled, tables FO, SS are just 1:1 links to the final data (in SA), in SA SAsampleWeightMeasured is enroll.

```{r}
apiclus2_H1 <- Pckg_survey_apiclus2_H1 # valid H1
apiclus2_H1
```

Six schools have no enrolment value (`SAsamp` is "N"). The survey package example `svytotal(~enroll, dclus2, na.rm = TRUE)` keeps these schools in the design with a value of zero, i.e. it estimates the total enrolment of the schools with a value. We do the same with `na = "zero"`. The dummy FO table claims 10 sampled units per trip while there is one, so we also need `strictSampleSize = FALSE`.

```{r}
apiclus2_H1$SA$SAsamp <- "Y"
res <- doEstimationForAllStrata(apiclus2_H1, "SAsampWtMes", na = "zero",
                                strictSampleSize = FALSE)
round(unlist(res[res$recType == "DE", c("est.total", "se.total")]), 1)
```

This is the result of the survey package: a total of 2,639,273 with a standard error of 799,638. The standard error includes the variance between the districts and the variance from sampling schools within the districts:

```{r}
round(unlist(res[res$recType == "VS", c("var.between", "var.within")]))
```

### Estimation from biological variables

For variables of the BV (or FM) table the design tree goes down to the fish. Here we estimate the numbers at age of sprat in the H8 example. The boxes (SA) are raised to the landing with the ratio of the total and sampled weights (`ratio = c(SA = "weight")`), see the ratio estimator vignette for more.

```{r}
tree <- createDesignTree(H8ExampleEE1, "BV", bvAssess = "Age",
                         ratio = c(SA = "weight"),
                         strictSampleSize = FALSE, varianceAsWR = "TE")
leaves <- getLeafValuesBV(tree, H8ExampleEE1, classMeas = "Age")
est <- doEstimationOnDesignTree(tree, leaves)
naa <- est$nodes[node == "ROOT"]
naa$Age <- as.integer(sub("N\\|Age=", "", naa$var))
naa <- naa[order(naa$Age), ]
knitr::kable(data.frame(Age = naa$Age, millions = naa$t / 1e6,
                        se = naa$se / 1e6),
             digits = 2,
             caption = "Numbers at age (millions) of the sampled months")
```

See also other package vignettes:

```{r,echo=F, results='asis', eval=T}
# note: vignettes are first compiled and only then html, R and Rmd are moved to folder doc
vignettes <- data.frame(Rmd=dir()[grepl(dir(), pat=".Rmd")])
# determines index of vignette being compiled
thisVignetteRmd<-paste0(gsub(":","",gsub(" ", "-",as.character(rmarkdown::metadata$title))),".Rmd")
# prints list with links
for (i in seq_along(vignettes$Rmd)) {
  if(tolower(vignettes$Rmd[i])!=tolower(thisVignetteRmd)){
  cat(sprintf("- [%s](%s)\n", gsub("-"," ",gsub(".Rmd","",vignettes$Rmd[i])), gsub(".Rmd",".html",vignettes$Rmd[i])))
}
}
```

#END

