---
title: "03b Estimating population parameters: ratio estimator"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{03b Estimating population parameters: ratio estimator}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


## Introduction

Catch numbers at age are usually estimated by raising the age composition of the samples to the landings: a ratio estimator with the landed weight as the auxiliary variable. This vignette estimates the numbers at age of sprat landed in the first and the fourth quarter in two ways, from the same samples and the same landings:

1. **Simple estimation**: the fish of all samples of a quarter are pooled and their age composition is raised to the landings, with **addCLtoLowerCS(...)** and **doBVestimCANUM(...)**.
2. **Design-based estimation**: every sampling stage is taken into account, with **createDesignTree(...)**, **doEstimationOnDesignTree(...)** and **doRatioEstimationToTotal(...)**.

The results are shown side by side, and the last part explains where and why they differ.

```{r}
library(RDBEScore)
library(data.table)
```

## The data and its sampling design

The example dataset resembles in many ways real data at ICES area 27.3.d.28.1 from the Estonian Baltic trawling fleet for 2022 sprat (*Sprattus sprattus*). The dataset has both total landings (CL) and commercial sampling data.

```{r}
H8ExampleEE1
```

The data has not gone through the RDBES upload and download procedure and contains invalid data types but also may have some other validity problems. However the fields for estimation should be present and valid.

```{r eval=FALSE, echo=TRUE}
#to check data types
validateRDBESDataObject(H8ExampleEE1, checkDataTypes = TRUE, strict = FALSE)
```

The design tree lists the sampling stages of the data:

```{r}
x <- H8ExampleEE1
tree <- createDesignTree(x, "BV", bvAssess = "Age", ratio = c(SA = "weight"),
                         strictSampleSize = FALSE, varianceAsWR = "TE")
tree[, .(units = .N, method = paste(unique(method), collapse = ", ")),
     by = table]
```

Weeks (TE) are sampled within months (2 or 3 of the 4 weeks), vessels (VS) within the weeks, landings (LE) from the vessels and one box of sprat (SA) from each landing; all fish of a box are aged (BV). In two weeks there are two vessels although `VSnumSamp` is 1: `strictSampleSize = FALSE` keeps both, each raised as if it were the only one. The arguments `ratio` and `varianceAsWR` are explained below.

## Simple estimation

The samples and the landings have to be matched to the same strata: the area, the gear, the species and the months of the quarter. **addCLtoLowerCS(...)** extracts the BV data of the samples (lower hierarchy C) and adds the sum of the landings. **doBVestimCANUM(...)** pools all fish, computes the proportion at age and the mean weight and raises them to the landings $L$: the number of fish at age $a$ is $N_a = L \cdot n_a / w$, where $n_a$ is the number of sampled fish of age $a$ and $w$ the weight of all sampled fish.

```{r}
quarters <- list(Q1 = 1:3, Q4 = 10:12)
biolCL <- lapply(names(quarters), function(q) {
  strataListCS <- list(LEarea = "27.3.d.28.1",
                       LEmetier6 = "OTM_SPF_16-31_0_0",
                       TEstratumName = month.name[quarters[[q]]],
                       SAspeCodeFAO = "SPR")
  strataListCL <- list(CLarea = "27.3.d.28.1",
                       CLquar = as.integer(sub("Q", "", q)),
                       CLmetier6 = "OTM_SPF_16-31_0_0",
                       CLspecFAO = "SPR")
  # the warnings only say that the months are combined into one stratum
  suppressWarnings(
    addCLtoLowerCS(x, strataListCS, strataListCL, combineStrata = TRUE,
                   lowerHierarchy = "C", CLfields = "CLoffWeight"))
})
names(biolCL) <- names(quarters)

simple <- rbindlist(lapply(names(quarters), function(q) {
  # ages 0 to 7 and a plus group 8+
  r <- doBVestimCANUM(biolCL[[q]], "sumCLoffWeight", classUnits = "Ageyear",
                      classBreaks = 0:8)
  r[, .(quarter = q, Age = as.character(Group), simple = totNum)]
}))
```

The class breaks also decide which fish are used: with `classBreaks = 1:8` the fish of age 0 would be left out of the proportions at age.

## Design-based estimation

The design-based estimation follows the sampling stages of the design tree and uses two ratio estimators:

- **A ratio estimator at the box stage.** The fish of a box are raised to the landing it was taken from with the ratio of the total and the sampled weight, `SAtotalWtMes / SAsampWtMes` (`ratio = c(SA = "weight")`), instead of the number of boxes, `SAnumTotal / SAnumSamp`. The landings, vessels and weeks are raised by their selection probabilities.
- **A ratio estimator to the landings.** The numbers at age and the landed weight are estimated from the same samples. **doRatioEstimationToTotal(...)** then scales the numbers at age so that the estimated landings equal the CL landings of the quarter.

Most weeks have one sampled vessel, so the variance within the weeks cannot be estimated. `varianceAsWR = "TE"` approximates the variance by treating the weeks as sampled with replacement, which overestimates it somewhat.

```{r}
# the number of fish by age (8 and older as 8+) and the sampled weight
fish <- getLeafValuesBV(tree, x, classMeas = "Age")
age <- as.integer(sub("N|Age=", "", fish$var, fixed = TRUE))
fish[, var := paste0("N|Age=", ifelse(age >= 8, "8+", age))]
leaves <- rbind(fish, getLeafValuesSA(tree, x, "SAsampWtMes"))
est <- doEstimationOnDesignTree(tree, leaves)

# the quarter of each sample, from the month (TE stratum)
quarterOf <- function(nodes) {
  month <- getAncestorValue(tree, nodes, x, "TEstratumName")
  paste0("Q", ceiling(match(month, month.name) / 3))
}
# the CL landings, in grammes like SAsampWtMes
landings <- x$CL[CLarea == "27.3.d.28.1" &
                   CLmetier6 == "OTM_SPF_16-31_0_0" & CLspecFAO == "SPR",
                 .(L = sum(CLoffWeight) * 1000),
                 by = .(quarter = paste0("Q", CLquar))]
design <- doRatioEstimationToTotal(est, unique(fish$var), "SAsampWtMes",
                                   X = setNames(landings$L, landings$quarter),
                                   domainOf = quarterOf)
design <- design[, .(quarter = domain,
                     Age = sub("N|Age=", "", var, fixed = TRUE),
                     design = est, se)]
```

## The results side by side

```{r}
cmp <- merge(simple[, .(quarter, Age, simple)], design,
             by = c("quarter", "Age"), all = TRUE)
cmp[is.na(simple), simple := 0]
# ages found in neither estimate (age 0 in the first quarter)
cmp <- cmp[simple > 0 | design > 0]
cmp[, difference := fifelse(simple > 0, 100 * (design / simple - 1),
                            NA_real_)]
cmp[, ageOrder := as.integer(sub("+", "", Age, fixed = TRUE))]
setorder(cmp, quarter, ageOrder)
knitr::kable(cmp[, .(Quarter = quarter, Age,
                     `Simple (millions)` = simple / 1e6,
                     `Design-based (millions)` = design / 1e6,
                     `Difference (%)` = difference,
                     `Design-based CV (%)` = 100 * se / design)],
             digits = 2,
             caption = "Numbers at age of sprat raised to the CL landings")
```

```{r}
totals <- cmp[, .(simple = sum(simple) / 1e6, design = sum(design) / 1e6),
              by = quarter]
totals
```

The total numbers are close (`r round(totals$simple[1], 1)` and `r round(totals$design[1], 1)` million in the first quarter), but the age compositions differ: the numbers of single ages differ by up to `r round(max(abs(cmp$difference), na.rm = TRUE))`%. Only the design-based estimates come with standard errors.

### Numbers at length

The same comparison for 10 mm length classes (`classBreaks = seq(70, 130, 10)`, with a plus group 130+ mm). For the design-based estimate the fish units are those measured for length (`bvAssess = "LengthTotal"`).

```{r}
simpleL <- rbindlist(lapply(names(quarters), function(q) {
  r <- doBVestimCANUM(biolCL[[q]], "sumCLoffWeight", classUnits = "Lengthmm",
                      classBreaks = seq(70, 130, 10))
  r[, .(quarter = q, Length = as.character(Group), simple = totNum)]
}))

treeL <- createDesignTree(x, "BV", bvAssess = "LengthTotal",
                          ratio = c(SA = "weight"),
                          strictSampleSize = FALSE, varianceAsWR = "TE")
fishL <- getLeafValuesBV(treeL, x, classMeas = "LengthTotal",
                         breaks = c(seq(70, 130, 10), Inf))
estL <- doEstimationOnDesignTree(treeL, rbind(fishL,
                                              getLeafValuesSA(treeL, x,
                                                              "SAsampWtMes")))
designL <- doRatioEstimationToTotal(estL, unique(fishL$var), "SAsampWtMes",
                                    X = setNames(landings$L, landings$quarter),
                                    domainOf = quarterOf)
# class "[70,80)" is "70-80" and "[130,Inf)" is "130+" in doBVestimCANUM
classLabel <- sub("N|LengthTotal=", "", designL$var, fixed = TRUE)
bounds <- tstrsplit(gsub("[][()]", "", classLabel), ",")
designL[, Length := fifelse(bounds[[2]] == "Inf", paste0(bounds[[1]], "+"),
                            paste0(bounds[[1]], "-", bounds[[2]]))]

cmpL <- merge(simpleL, designL[, .(quarter = domain, Length, design = est, se)],
              by = c("quarter", "Length"), all = TRUE)
cmpL[is.na(simple), simple := 0]
cmpL[, lengthOrder := as.numeric(sub("[-+].*$", "", Length))]
setorder(cmpL, quarter, lengthOrder)
knitr::kable(cmpL[, .(Quarter = quarter, `Length (mm)` = Length,
                      `Simple (millions)` = simple / 1e6,
                      `Design-based (millions)` = design / 1e6,
                      `Difference (%)` = fifelse(simple > 0,
                                                 100 * (design / simple - 1),
                                                 NA_real_),
                      `Design-based CV (%)` = 100 * se / design)],
             digits = 2,
             caption = "Numbers at length of sprat raised to the CL landings")
```

The differences have the same cause as for the age compositions, explained next.

## Why the estimates differ

Both estimators raise the samples $k$ of a quarter to the same landings $L$, and both can be written as

$$N_a = L \sum_k s_k \frac{n_{a,k}}{w_k}$$

where $n_{a,k} / w_k$ is the number of fish of age $a$ per gram in sample $k$. They differ only in the weight $s_k$ that each sample gets:

- **Simple**: $s_k = w_k / \sum_j w_j$, the share of the sample in the sampled weight. A sample of 1 kg counts twice as much as a sample of 0.5 kg, whatever the size of the landing it comes from.
- **Design-based**: $s_k = W_k X_k / \sum_j W_j X_j$, the share of the landings the sample represents: the weight $X_k$ (`SAtotalWtMes`) of the landing it comes from, raised by $W_k$ for the selection of the landings, the vessels and the weeks.

The weights of the samples are:

```{r}
# raising factor of a unit: product of the weights of it and the units above
raising <- function(nodes) {
  w <- rep(1, length(nodes))
  repeat {
    i <- match(nodes, tree$node)
    if (all(is.na(i))) break
    w <- w * fifelse(is.na(i), 1, tree$w[i])
    nodes <- tree$parent[i]
  }
  w
}
samples <- tree[table == "SA", .(node, SAid = as.numeric(sub("SA:", "", node)),
                                 W = raising(parent))]
samples <- x$SA[, .(SAid, sampled = SAsampWtMes, landing = SAtotalWtMes)][
  samples, on = "SAid"]
samples[, quarter := quarterOf(samples$node)]
samples[, month := getAncestorValue(tree, samples$node, x, "TEstratumName")]
samples[, `:=`(simpleShare = 100 * sampled / sum(sampled),
               designShare = 100 * W * landing / sum(W * landing)),
        by = quarter]
setorder(samples, quarter, SAid)
knitr::kable(samples[, .(Quarter = quarter, Month = month, Sample = SAid,
                         `Sampled (kg)` = sampled / 1000,
                         `Landing (t)` = landing / 1e6,
                         `Raising of the landing` = W,
                         `Landings represented (t)` = W * landing / 1e6,
                         `Simple weight (%)` = simpleShare,
                         `Design-based weight (%)` = designShare)],
             digits = 1, caption = "The weight of each sample")
```

```{r, include = FALSE}
q1 <- samples[quarter == "Q1"]
over <- q1[which.max(simpleShare - designShare)]
under <- q1[which.max(designShare - simpleShare)]
```

In the first quarter sample `r over$SAid` (`r over$month`) has `r round(over$simpleShare)`% of the sampled weight but represents only `r round(over$designShare)`% of the landings, while sample `r under$SAid` (`r under$month`) has `r round(under$simpleShare)`% of the sampled weight and represents `r round(under$designShare)`% of the landings. The simple estimator gives the age composition of the first sample far too much weight and that of the second far too little.

The formula reproduces both estimates (in the simple estimator $w_k$ is the weight of the aged fish, here within 0.2% of `SAsampWtMes`):

```{r}
# number and weight of the aged fish of each sample
bv <- dcast(x$BV[BVtypeMeas %in% c("Age", "WeightMeasured")],
            SAid + BVfishId ~ BVtypeMeas, value.var = "BVvalueMeas")
bv[, Age := fifelse(as.integer(Age) >= 8, "8+", Age)]
perSample <- bv[, .(fishWeight = sum(as.numeric(WeightMeasured))),
                by = SAid][samples, on = "SAid"]
perSample[, fishShare := fishWeight / sum(fishWeight), by = quarter]
perAge <- perSample[bv[, .(n = .N), by = .(SAid, Age)], on = "SAid"]
formula <- perAge[, .(simple = sum(fishShare * n / fishWeight),
                      design = sum(designShare / 100 * n / sampled)),
                  by = .(quarter, Age)]
formula <- landings[formula, on = "quarter"]
formula[, `:=`(simple = L * simple, design = L * design)]
check <- cmp[formula, on = .(quarter, Age)]
c(simple = isTRUE(all.equal(check$i.simple, check$simple)),
  design = isTRUE(all.equal(check$i.design, check$design)))
```

The two estimators agree only when each sample represents the same share of the landings as its share of the sampled weight, that is when the samples are taken in proportion to the landings. Here the weeks, vessels and landings were sampled with different probabilities and the boxes come from landings of very different size, so the simple estimator over-represents some samples and under-represents others.

Both estimators share the limits of the sampling: the second and third quarter have landings but no samples and cannot be raised, and in the fourth quarter the age composition of October and November is applied to the December landings as well. The estimated landings show how much the ratio to the CL landings corrects:

```{r}
landingsHat <- doRatioEstimationToTotal(est, "SAsampWtMes", "SAsampWtMes",
                                        X = setNames(landings$L,
                                                     landings$quarter),
                                        domainOf = quarterOf)
setorder(landingsHat, domain)
knitr::kable(landingsHat[, .(Quarter = domain, `CL landings (t)` = X / 1e6,
                             `Estimated from the samples (t)` = t_x / 1e6)],
             digits = 1)
```

## Exporting results to InterCatch

The results can be formatted for the InterCatch data call using **exportEstimationResultsToInterCatchFormat()**. Here we export the design-based numbers at age with their variances. The mean weight at age is the ratio of the estimated weight and number of fish at age in each quarter.

```{r}
# number and weight of fish by age (8+) and quarter
byQuarter <- rbind(getLeafValuesBV(tree, x, classMeas = "Age"),
                   getLeafValuesBV(tree, x, classMeas = "Age",
                                   sumMeas = "WeightMeasured"))
parts <- tstrsplit(byQuarter$var, "|Age=", fixed = TRUE)
ageQ <- as.integer(parts[[2]])
byQuarter[, var := paste(parts[[1]],
                         fifelse(ageQ >= 8, "8+", as.character(ageQ)),
                         quarterOf(node), sep = "|")]
estQ <- doEstimationOnDesignTree(tree, byQuarter)$nodes[node == "ROOT"]
estQ[, c("measure", "Age", "quarter") := tstrsplit(var, "|", fixed = TRUE)]
meanWeight <- dcast(estQ, quarter + Age ~ measure, value.var = "t")
meanWeight[, MeanWeight := WeightMeasured / N]
results <- meanWeight[design, on = .(quarter, Age)][design > 0]
setorder(results, quarter, Age)

dataToExport <- data.frame(
    # HI columns
    Country = "EE",
    Year = "2022",
    SeasonType = "Quarter",
    Season = as.integer(sub("Q", "", results$quarter)),
    Fleet = "OTM_SPF_16-31_0_0",
    AreaType = "Div",
    FishingArea = "27.3.d.28.1",
    DepthRange = "NA",

    # SI columns
    Species = "SPR",
    Stock = "Unknown",
    CatchCategory = "L",
    ReportingCategory = "R",
    DataToFrom = "Data",
    Usage = "Used",
    SamplesOrigin = "U",
    QualityFlag = "U",
    UnitCATON = "kg",
    CATON = landings$L[match(results$quarter, landings$quarter)] / 1000,
    OffLandings = landings$L[match(results$quarter, landings$quarter)] / 1000,

    # SD columns
    Sex = "N",
    CANUMtype = "Age",
    AgeLength = as.integer(sub("+", "", results$Age, fixed = TRUE)),
    PlusGroup = 8,
    unitMeanWeight = "g",
    unitCANUM = "n",
    UnitAgeOrLength = "year",
    NumberCaught = results$design,
    MeanWeight = results$MeanWeight,
    varNumLanded = results$se^2
)

# Generate the InterCatch string
ic_output <- exportEstimationResultsToInterCatchFormat(dataToExport)

# Show the first few lines of the output
head(ic_output)
```


See also other package vignettes:

```{r,echo=F, results='asis', eval=T}
# note: vignettes are first compiled and only then html, R and Rmd are moved to folder doc
vignettes <- data.frame(Rmd=dir()[grepl(dir(), pat=".Rmd")])
# determines index of vignette being compiled
thisVignetteRmd<-paste0(gsub(":","",gsub(" ", "-",as.character(rmarkdown::metadata$title))),".Rmd")
# prints list with links
for (i in seq_along(vignettes$Rmd)) {
  if(tolower(vignettes$Rmd[i])!=tolower(thisVignetteRmd)){
  cat(sprintf("- [%s](%s)\n", gsub("-"," ",gsub(".Rmd","",vignettes$Rmd[i])), gsub(".Rmd",".html",vignettes$Rmd[i])))
}
}
```

#END
